---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# summarystatistics descrtiption  

The R package `summarystatistics` includes the function `summarize_stats`. The function computes five summary statistics (minimum, maximum, mean, and two quantiles) of a numerical variable across the groups of a categorical variable. The function also produces simple boxplots that allow for the visual comparison of distributions between the groups of the categorical variable.


## Installation instructions  

To install `summarystatistics` package from GitHub use:

``` r
devtools::install_github("kearutherford/Assignment_B2_Rutherford")
```


## Demonstrations  

```{r}
# Load required packages -----------------------------------------------------------------------------
suppressPackageStartupMessages(library(datateachr))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(palmerpenguins))
#suppressPackageStartupMessages(library(dplyr))
#suppressPackageStartupMessages(library(ggplot2))
#suppressPackageStartupMessages(library(magrittr))
#suppressPackageStartupMessages(library(stats))
#suppressPackageStartupMessages(library(tibble))
#suppressPackageStartupMessages(library(tidyr))
#suppressPackageStartupMessages(library(testthat))
suppressPackageStartupMessages(library(summarystatistics))
```

The following examples demonstrate how to use the `summarize_stats` function in the R package `summarystatistics`. The examples rely on the [vancouver_trees](https://opendata.vancouver.ca/explore/dataset/street-trees/information/?disjunctive.species_name&disjunctive.common_name&disjunctive.height_range_id&location=11,49.26339,-123.12046) data set included in the `datateachr` package, the [gapminder](https://www.rdocumentation.org/packages/gapminder/versions/0.3.0) data set included in the `gapminder` package, and the [penguins](https://allisonhorst.github.io/palmerpenguins/) data set included in the `palmerpenguins` package. 

### Example #1: Character variable input, keep default arguments

**Description of inputs:**  
This example demonstrates inputting a variable with a character class for the `categoric_var` argument. The default arguments are kept (`prob1` = 0.25, `prob2` = 0.75, `na.rm` = TRUE).  

|Argument|Input|Description of Input|Class|
|:---|:---|:---|:---|
|data|vancouver_trees|provides an inventory of public street trees in Vancouver|tbl_df|
|categoric_var|root_barrier|root_barrier installed (Y = Yes, N = No)|character|
|numeric_var|diameter|diameter at breast height (DBH) of trees in inches|numeric|  

```{r}
# Run summarize_stats() function with vancouver_trees data set ---------------------------------------
summarize_stats(data = datateachr::vancouver_trees, 
                categoric_var = root_barrier,
                numeric_var = diameter)
```

**Description of output:** The output is a list with two items: a tibble of summary statistics and a simple set of boxplots. For the summary tibble, the minimum, maximum, mean, first quartile, and third quartile were calculated for DBH across the two root barrier groups (Y/N). The boxplots show the difference in the distributions of DBH between the two root barrier groups.  

### Example #2: Factor variable input, alter default arguments  

**Description of inputs:**  
This example demonstrates inputting a variable with a factor class for the `categoric_var` argument. The ` prob1` and `prob2` arguments are set to `prob1` = 0.1 and `prob2` = 0.9. The default argument `na.rm = TRUE` is kept.  

|Argument|Input|Description of Input|Class|
|:---|:---|:---|:---|
|data|gapminder|contains global socioeconomic information|tbl_df|
|categoric_var|continent|five continents (Africa, Americas, Asia, Europe, Oceania)|factor|
|numeric_var|lifeExp|life expectancy at birth|numeric|  

```{r}
# Run summarize_stats() function with gapminder data set ---------------------------------------------
summarize_stats(data = gapminder::gapminder, 
                categoric_var = continent,
                numeric_var = lifeExp,
                prob1 = 0.1,
                prob2 = 0.9)
```

**Description of output:** The output is a list with two items: a tibble of summary statistics and a simple boxplot. For the summary tibble, the minimum, maximum, mean, 10th percentile, and 90th percentile were calculated for life expectancy across the five continents (Africa, Americas, Asia, Europe, and Oceania). The boxplots show the distribution of life expectancy for each continent.

### Example #3: Factor variable input, alter default arguments  

**Description of inputs:**  
This example also demonstrates inputting a variable with a factor class for the `categoric_var` argument. The ` prob1` and `prob2` arguments are set to `prob1` = 0.2 and `prob2` = 0.8. The default argument `na.rm = TRUE` is kept.  

|Argument|Input|Description of Input|Class|
|:---|:---|:---|:---|
|data|penguins|contains size measurements for penguins in Antarctica|tbl_df|
|categoric_var|species|three penguin species (Adelie, Chinstrap, Gentoo)|factor|
|numeric_var|flipper_length_mm|flipper length in millimeters|numeric|  

```{r}
# Run summarize_stats() function with penguins data set ----------------------------------------------
summarize_stats(data = palmerpenguins::penguins, 
                categoric_var = species,
                numeric_var = flipper_length_mm,
                prob1 = 0.2,
                prob2 = 0.8)
```

**Description of output:** The output is a list with two items: a tibble of summary statistics and a simple boxplot. For the summary tibble, the minimum, maximum, mean, 20th percentile, and 80th percentile were calculated for flipper length across the three species (Adelie, Chinstrap, and Gentoo). The boxplots show the distribution of flipper length for each species.  

### Example #4: Error from invalid argument  

**Description of inputs:**  
This example demonstrates inputting a factor variable for the `numeric_var` argument.  

|Argument|Input|Description of Input|Class|
|:---|:---|:---|:---|
|data|penguins|contains size measurements for penguins in Antarctica|tbl_df|
|categoric_var|species|three penguin species (Adelie, Chinstrap, Gentoo)|character|
|numeric_var|island|three islands (Biscoe, Dream, Torgersen)|factor|  

```{r, error = TRUE}
# Run summarize_stats() function with penguins data set ----------------------------------------------
summarize_stats(data = palmerpenguins::penguins, 
                categoric_var = species,
                numeric_var = island)
```

**Description of output:** The function throws an error message. The message informs the user that the parameter `numeric_var` requires the input of a numerical variable. In this case the input for the argument was a variable of class factor. Custom error messages also appear for invalid arguments of the other two parameters (`data` and `categoric_var`).  
