---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# summarystatistics  

The R package `summarystatistics` includes the function `summarize_stats`. The function computes five summary statistics (minimum, maximum, mean, and two quantiles) of a numerical variable across the groups of a categorical variable. The function also produces a simple boxplot that allows for visual comparison between the distributions of the numerical variable across the different groups of the categorical variable.

## Installation instructions  

Install `summarystatistics` package from GitHub:

``` r
devtools::install_github("kearutherford/Assignment_B2_Rutherford")
```

## Demonstrations  

```{r}
# Load required packages -------------------------------------------------------

suppressPackageStartupMessages(library(datateachr))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(palmerpenguins))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(tibble))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(summarystatistics))
```

The following examples demonstrate how to use the `summarize_stats` function in the R package `summarystatistics`. The examples rely on the [vancouver_trees](https://opendata.vancouver.ca/explore/dataset/street-trees/information/?disjunctive.species_name&disjunctive.common_name&disjunctive.height_range_id&location=11,49.26339,-123.12046) data set included in the `datateachr` package, the [gapminder](https://www.rdocumentation.org/packages/gapminder/versions/0.3.0) data set included in the `gapminder` package, and the [penguins](https://allisonhorst.github.io/palmerpenguins/) data set included in the `palmerpenguins` package. 

### Example #1: Character variable input, keep default arguments

**Description of inputs:** This first example demonstrates inputting a variable with a character class for the `categoric_var` argument. The default arguments are kept (`prob1` = 0.25, `prob2` = 0.75, `na.rm` = TRUE).  

|Argument|Input|Description|Class|
|:---|:---|:---|:---|
|data|vancouver_trees|the data set provides an inventory of the public street trees in Vancouver, BC|tbl_df|
|categoric_var|root_barrier|root_barrier installed (Y = Yes, N = No)|character|
|numeric_var|diameter|diameter at breast height (DBH) of trees in inches|numeric|  

```{r}
# Run summarize_stats() function with Van_trees_2 data set ---------------------
summarize_stats(data = datateachr::vancouver_trees, 
                categoric_var = root_barrier,
                numeric_var = diameter)
```

**Description of output:** The output is a list with two items: a tibble of summary statistics and a simple boxplot. For the summary tibble, the minimum, maximum, mean, first quartile, and third quartile were calculated for DBH across the two root barrier groups (Y/N). The boxplot shows the difference in the distributions of DBH between the two root barrier groups.  


### Example #2: Factor variable input, alter default arguments  

**Description of inputs:** This second example demonstrates inputting a variable with a factor class for the `categoric_var` argument. The ` prob1` and `prob2` default arguments are altered to `prob1` = 0.1 and `prob2` = 0.9. The default argument `na.rm = TRUE` is kept.  

|Argument|Input|Description|Class|
|:---|:---|:---|:---|
|data|gapminder|the data set provides global socioeconomic information|tbl_df|
|categoric_var|continent|five continents|factor|
|numeric_var|lifeExp|life expectancy at birth|numeric|  

```{r}
# Run summarize_stats() function with van_trees_2 data set ---------------------
summarize_stats(data = gapminder::gapminder, 
                categoric_var = continent,
                numeric_var = lifeExp,
                prob1 = 0.1,
                prob2 = 0.9)
```

**Description of output:** The output is a list with two items: a tibble of summary statistics and a simple boxplot. For the summary tibble, the minimum, maximum, mean, 10th percentile, and 90th percentile were calculated for life expectancy across the five continents (Africa, Americas, Asia, Europe, and Oceania). The boxplot shows the distribution of life expectancy for each continent.  


### Example #3: Factor variable input, alter default arguments  

**Description of inputs:** This third example also demonstrates inputting a variable with a factor class for the `categoric_var` argument. The ` prob1` and `prob2` default arguments are altered to `prob1` = 0.2 and `prob2` = 0.8. The default argument `na.rm = TRUE` is kept.  

|Argument|Input|Description|Class|
|:---|:---|:---|
|data|penguins|the data set provides size measurements for penguins in Antarctica|tbl_df|
|categoric_var|species|three penguin species|factor|
|numeric_var|flipper_length_mm|flipper length in mm|numeric|  

```{r}
# Run summarize_stats() function with van_trees_2 data set ---------------------
summarize_stats(data = palmerpenguins::penguins, 
                categoric_var = species,
                numeric_var = flipper_length_mm,
                prob1 = 0.2,
                prob2 = 0.8)
```
**Description of output:** The output is a list with two items: a tibble of summary statistics and a simple boxplot. For the summary tibble, the minimum, maximum, mean, 20th percentile, and 90th percentile were calculated for flipper length across the three species (Adelie, Chinstrap, and Gentoo). The boxplot shows the distribution of flipper length for each species.  


### Example #4: Error from invalid argument  

**Description of inputs:** This forth example demonstrates inputting a factor variable for the `numeric_var` argument.  

|Argument|Input|Description|Class|
|:---|:---|:---|
|data|penguins|the data set provides size measurements for penguins in Antarctica|tbl_df|
|categoric_var|species|three penguin species|character|
|numeric_var|island|three islands|factor|  

```{r, error = TRUE}
# Run summarize_stats() function with van_trees_2 data set --------------------------------------------------------------------------------
summarize_stats(data = palmerpenguins::penguins, 
                categoric_var = species,
                numeric_var = island)
```

**Description of output:** The function throws an error message. The message informs the user that the parameter `numeric_var` requires the input of a numerical variable. In this case the input for the argument was a variable of class factor. Custom error messages also appear for invalid arguments of the other two parameters (`data` and `categoric_var`).  
